Ancestry inference and grouping\\from principal component analysis of genetic data

### 2020-09-26 paper/paper-ancestry-matching.tex
start writing paper again

\title{Insights on ancestry using principal component analysis of genetic data}
\author{Florian Priv\'e$^{\text{1,}*}$ and Bjarni J. Vilhj\'almsson$^{\text{1}}$}

@@

\section{Introduction}

@@

\section{Material and Methods}

@@

\subsection{Distance between two populations}

To infer a PCA-based genetic distance between two populations, we use the Bhattacharyya distance between two multivariate normal distributions $\mathcal{N}(\boldsymbol\mu_1,\,\boldsymbol\Sigma_1)$ and $\mathcal{N}(\boldsymbol\mu_2,\,\boldsymbol\Sigma_2)$ as defined as
$$D_B={1\over 8}(\boldsymbol\mu_2-\boldsymbol\mu_1)^T \boldsymbol\Sigma^{-1}(\boldsymbol\mu_2-\boldsymbol\mu_1)+{1\over 2}\log \,\left({|\boldsymbol\Sigma| \over \sqrt{|\boldsymbol\Sigma_1| \, |\boldsymbol\Sigma_2|} }\right)~,$$
where $\boldsymbol\Sigma={\boldsymbol\Sigma_1+\boldsymbol\Sigma_2 \over 2}$ and $|M|$ is the absolute value of the determinant of matrix $M$ \cite[]{bhattacharyya1943measure,fukunaga1990introduction}. 
The mean and covariance parameters for each population are computed using the robust center and covariance parameters as in section \ref{ancestry}.

@@

\noindent[DISCUSS LACK OF CERTAIN POPULATIONS IN 1000G]

@@

F.P.\ and B.V.\ are supported by the Danish National Research Foundation (Niels Bohr Professorship to John McGrath).

>>>>

\title{Ancestry inference and grouping\\from principal component analysis of genetic data
}
\author{Florian Priv\'e$^{\text{1,}*}$}

@@

$^\dag$ Further defined in supplementary section ``Defintions''.

\section*{Introduction}

Why do we need to group individuals in ancestry groups?

There is no consensus on the method to use to infer population structure, but particularly on grouping individuals.

Why it makes sense to use PCA-based distances -> correction for population structure, and geographical distance.

@@

\section*{Measures of genetic dissimilarity between populations}

We first compare four measures of genetic dissimilarity using populations of the 1000 genomes as example \cite[]{10002015global}. 
The $F_{ST}$$^\dag$ is an ubiquitous measure of genetic dissimilarity between populations and the first measure we use in this comparison.
The other three measures compared are distances applied to the PC scores$^\dag$ of the genetic data: 1) the Bhattacharyya distance$^\dag$; 2) the distance between the centers (geometric median$^\dag$) of the two populations; 3) the shorter distance between pairs of PC scores from the two populations.
The distance between population centers seems to be an appropriate PCA-based distance as this (squared) distance is approximately proportional to the $F_{ST}$ (Figure [TODO]) and provides an appropriate clustering of populations (Figure [TODO]).
In contrast, the two other Bhattacharyya and shortest distances do not provide as satisfactory results (Figures [TODO]).

%\section{Material and Methods}

\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}
\renewcommand{\theequation}{S\arabic{equation}}
\setcounter{equation}{0}
\renewcommand{\thesection}{S\arabic{section}}
\setcounter{section}{1}

\section*{Definitions $\dag$}

\begin{itemize}

	\item The $\boldsymbol{F_{ST}}$ measures the relative amount of genetic variance between populations compared to the total genetic variance within these populations \cite[]{wright1965interpretation}.
	We use the weighted average formula proposed in \cite{weir1984estimating}, which we now implement in our package bigsnpr \cite[]{prive2017efficient}.

	\item The {\bf Principal Component (PC) scores} are defined as $U \Delta$, where $U \Delta V^T$ is the singular value decomposition of the (scaled) genotype matrix. They are usually truncated, e.g.\ corresponding to the first 20 principal dimensions only.

	\item The {\bf Bhattacharyya distance} between two multivariate normal distributions $\mathcal{N}(\boldsymbol\mu_1,\,\boldsymbol\Sigma_1)$ and $\mathcal{N}(\boldsymbol\mu_2,\,\boldsymbol\Sigma_2)$, which is defined as
	$D_B={1\over 8}(\boldsymbol\mu_2-\boldsymbol\mu_1)^T \boldsymbol\Sigma^{-1}(\boldsymbol\mu_2-\boldsymbol\mu_1)+{1\over 2}\log \,\left({|\boldsymbol\Sigma| \over \sqrt{|\boldsymbol\Sigma_1| \, |\boldsymbol\Sigma_2|} }\right)$,
	where $\boldsymbol\Sigma={\boldsymbol\Sigma_1+\boldsymbol\Sigma_2 \over 2}$ and $|M|$ is the absolute value of the determinant of matrix $M$ \cite[]{bhattacharyya1943measure,fukunaga1990introduction}. 
	The mean and covariance parameters for each population are computed using the robust location and covariance parameters as proposed in \cite{prive2020efficient}.

	\item The {\bf geometric median} of points is the point that minimizes the sum of all Euclidean distances to these points. We now implement this as function \texttt{geometric\_median} in our R package bigutilsr.

\end{itemize}

@@

@@

\noindent[DISCUSS LACK OF SOME POPULATIONS IN 1000G]

@@

\section*{Funding}

F.P.\ is supported by the Danish National Research Foundation (Niels Bohr Professorship to John McGrath).
